---
title: "transient mass balance"
author: "Abby Mauger"
date: "2023-11-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
library(patchwork)
```

Using the first-order exponential integrator, we have

$$ C_{t+1} = E/Q\{1-exp(-Q/V\Delta t)\} + (C_{t}-C_R)exp(-Q/V\Delta t) +C_R $$
$ E = n * CO2rate $ is the CO2 emission rate in ug/s, $Q$ is the ventilation rate in m3/s, $V$ is the volume of the room in m3, $\Delta t$ is the time step in seconds, $C_R$ is the environmental CO2 concentration in ug/m3, and $C_t$ is the CO2 concentration at time $t$ in ug/m3.

We want to minimize the sum of squares between the predicted CO2 concentration and observed CO2 concentration:

$$ SS = \sum_{i=1}^{n} (\hat{C_{i}} - C_{i})^2 = \sum_{i=1}^{n}(E/Q\{1-exp(-Q/V\Delta t)\} + (C_{i-1}-C_R)exp(-Q/V\Delta t) +C_R) -C_i)^2$$

If $E$, $C$, and $C_R$ are all measured, we can use the Newton-Raphson method to find the value of $Q$ that minimizes the sum of squares.

Each update step of the Newton-Raphson algorithm will be:

$$Q_{t+1} = Q_t - \frac{\frac{\partial SS}{\partial Q}}{\frac{\partial^2SS}{\partial Q^2}}$$


Taking the first derivative with respect to Q we get:

$$ \frac{\partial SS}{\partial Q} = \sum_{i=1}^{n} 2(\hat{C_i}-C_i)[(-E/Q^2)(1-exp^{-Q/V\Delta t})+(E/Q + C_R-C_{i-1})
exp(-Q/V\Delta t)(\Delta t/V)]$$

Define the following:

$$ X = \hat{C_i}-C_i $$

$$ X' = \frac{\partial X}{\partial Q} = (-E/Q^2)(1-exp^{-Q/V\Delta t})+(E/Q + C_R-C_{i-1}) exp(-Q/V\Delta t)\Delta t/V $$
Then the second derivative with respect to Q is:

$$ \frac{\partial^2SS}{\partial Q^2} = 2\sum_{i=1}^{n} X'^2 + X[2E/Q^3(1-exp(-Q/V \Delta t)) -2E/Q^2exp(-Q/V \Delta t)\Delta t/V - (E/Q + C_R - C_{i-1})exp(-Q/V \Delta t)(\Delta t/V)^2] $$

Let's evaluate the NR method on simulated datasets.

```{r}
# simulated data for testing
# consider a 24 hour time period, starting at midnight and running until the following midnight
# We have 17 people enter at 8 am, leave for one hour at noon, and leave for the day at 5 pm
# 2 adults and 15 children 
persondata <- data.frame(
  time = c(0, 8, 8, 12, 12, 13, 13, 17, 24),
  n = c(0, 15, 2, 0, 0, 15, 2, 0, 0),
  age = c(0, 8, 30, 8, 30, 8, 30, 0, 0),
  gender = rep(NA, 9),
  MET = c(NA, 3, 1.8, 3, 1.8, 3, 2, NA, NA),
  CO2rate = rep(NA, 9)
)

freq=1
volume=500
envCO2=400
temp=25

# function to simulate data then run NR method
testing_NR <- function(Q, var, init.Q){
  sim <- simulateData(persondata = persondata,
                        volume=volume, 
                        ventilation_rate = Q, 
                        CO2var = var,
                        method='Exponential', 
                        freq=freq)
  out <- estimate_ventilation(freq=freq, 
                     CO2=sim$CO2, 
                     volume=volume, 
                     envCO2known=envCO2, 
                     init.Q=init.Q, 
                     temp=temp, 
                     persondata=persondata, 
                     method='NR', 
                     max.iter=1000, 
                     tol=1e-10)
  return(c(Q.est = out$est_ventilation, iter = out$iter, convergence = out$convergence, Q.true = Q, var = var, Q.init = init.Q))
}

Q = rep(seq(0.01, 1, length=10), 5) # do 5 simulations on each Q
var = c(.01, .1, 1, 2)
test_cases = merge(Q, var)


out_results = matrix(NA, nrow = nrow(test_cases), ncol = 6)
colnames(out_results) = c("Q.est", "iter", "convergence", "Q.true", "var", "Q.init")
for(i in 1:nrow(test_cases)){
  #print(paste("Running test case", i, "of", nrow(test_cases), "test cases."))
  out_results[i,] = testing_NR(test_cases[i, 1], test_cases[i, 2], 1)
}
```


```{r}
out_results <- as.data.frame(out_results)
out_results <- out_results %>% mutate(rel.error = abs(Q.est-Q.true)/Q.true)
summary <- out_results %>% group_by(Q.true, var) %>% summarize(mean.Q.est = mean(Q.est), mean.iter = mean(iter), mean.convergence = mean(convergence), mean.rel.error = mean(rel.error), sd.rel.error = sd(rel.error))

ggplot(summary, aes(x=Q.true, y=mean.rel.error)) +     geom_point() + 
  geom_line() +  
  geom_ribbon(aes(ymin=mean.rel.error-sd.rel.error, ymax=mean.rel.error+sd.rel.error), alpha=.2) + facet_wrap(~paste('Simulated CO2 variance:',var))+
  labs(x="True Ventilation Rate (m3/s)", y="Relative error in estimated ventilation rate") + theme_bw() + ylim(-0.1, 1)
```

```{r}
# example of what the simulated data looks like
sim_example <- simulateData(persondata = persondata,
                        volume=volume, 
                        ventilation_rate = .1, 
                        CO2var = .01,
                        method='Exponential', 
                        freq=freq)
plot.01 <- ggplot(sim_example, aes(x=time, y=CO2)) + geom_line() + labs(x="Time (hours)", y="CO2 (ppm)") + theme_bw() + ggtitle('CO2 variance = 0.01') + geom_vline(xintercept=c(8, 12, 13, 17), col='red', lty=2)
sim_example <- simulateData(persondata = persondata,
                        volume=volume, 
                        ventilation_rate = .1, 
                        CO2var = .1,
                        method='Exponential', 
                        freq=freq)
plot.1 <- ggplot(sim_example, aes(x=time, y=CO2)) + geom_line() + labs(x="Time (hours)", y="CO2 (ppm)") + theme_bw() + ggtitle('CO2 variance = 0.1') + geom_vline(xintercept=c(8, 12, 13, 17), col='red', lty=2)
sim_example <- simulateData(persondata = persondata,
                        volume=volume, 
                        ventilation_rate = .1, 
                        CO2var = 1,
                        method='Exponential', 
                        freq=freq)
plot1 = ggplot(sim_example, aes(x=time, y=CO2)) + geom_line() + labs(x="Time (hours)", y="CO2 (ppm)") + theme_bw() + ggtitle('CO2 variance = 1') + geom_vline(xintercept=c(8, 12, 13, 17), col='red', lty=2)
sim_example <- simulateData(persondata = persondata,
                        volume=volume, 
                        ventilation_rate = .1, 
                        CO2var = 2,
                        method='Exponential', 
                        freq=freq)
plot10 = ggplot(sim_example, aes(x=time, y=CO2)) + geom_line() + labs(x="Time (hours)", y="CO2 (ppm)") + theme_bw() + ggtitle('CO2 variance = 2') + geom_vline(xintercept=c(8, 12, 13, 17), col='red', lty=2)
plot.01+plot.1+plot1+plot10

```

What if we don't observe the CO2 as frequently? We will still simulate a measurement every 1 second, but will only keep one per minute.

```{r}
# simulated data for testing
# consider a 24 hour time period, starting at midnight and running until the following midnight
# We have 17 people enter at 8 am, leave for one hour at noon, and leave for the day at 5 pm
# 2 adults and 15 children 
persondata <- data.frame(
  time = c(0, 8, 8, 12, 12, 13, 13, 17, 24),
  n = c(0, 15, 2, 0, 0, 15, 2, 0, 0),
  age = c(0, 8, 30, 8, 30, 8, 30, 0, 0),
  gender = rep(NA, 9),
  MET = c(NA, 3, 1.8, 3, 1.8, 3, 2, NA, NA),
  CO2rate = rep(NA, 9)
)

freq.sim=1
freq.est = 60
volume=500
envCO2=400
temp=25

# function to simulate data then run NR method
testing_NR <- function(Q, var, init.Q){
  sim <- simulateData(persondata = persondata,
                        volume=volume, 
                        ventilation_rate = Q, 
                        CO2var = var,
                        method='Exponential', 
                        freq=freq.sim)
  index = seq(1, length(sim$CO2), freq.est)
  CO2 = sim$CO2[index]
  out <- estimate_ventilation(freq=freq.est, 
                     CO2=CO2, 
                     volume=volume, 
                     envCO2known=envCO2, 
                     init.Q=init.Q, 
                     temp=temp, 
                     persondata=persondata, 
                     method='NR', 
                     max.iter=1000, 
                     tol=1e-10)
  return(c(Q.est = out$est_ventilation, iter = out$iter, convergence = out$convergence, Q.true = Q, var = var, Q.init = init.Q))
}

Q = rep(seq(0.01, 1, length=10), 5) # do 5 simulations on each Q
var = c(.01, .1, 1, 2)
test_cases = merge(Q, var)


out_results = matrix(NA, nrow = nrow(test_cases), ncol = 6)
colnames(out_results) = c("Q.est", "iter", "convergence", "Q.true", "var", "Q.init")
for(i in 1:nrow(test_cases)){
  #print(paste("Running test case", i, "of", nrow(test_cases), "test cases."))
  out_results[i,] = testing_NR(test_cases[i, 1], test_cases[i, 2], 1)
}

out_results <- as.data.frame(out_results)
out_results <- out_results %>% mutate(rel.error = abs(Q.est-Q.true)/Q.true)
summary <- out_results %>% group_by(Q.true, var) %>% summarize(mean.Q.est = mean(Q.est), mean.iter = mean(iter), mean.convergence = mean(convergence), mean.rel.error = mean(rel.error), sd.rel.error = sd(rel.error))

ggplot(summary, aes(x=Q.true, y=mean.rel.error)) +     geom_point() + 
  geom_line() +  
  geom_ribbon(aes(ymin=mean.rel.error-sd.rel.error, ymax=mean.rel.error+sd.rel.error), alpha=.2) + facet_wrap(~paste('Simulated CO2 variance:',var))+
  labs(x="True Ventilation Rate (m3/s)", y="Relative error in estimated ventilation rate") + theme_bw() + ylim(-0.1, 1)
```

```{r}
# benchmarking data simulation
library(bench)
Cppresults <- bench::mark(simulateData(persondata = persondata,
                        volume=volume, 
                        ventilation_rate = .1, 
                        CO2var = 2,
                        method='Exponential', 
                        freq=freq), iterations = 100)
Rresults <- bench::mark(simulateData(persondata = persondata,
                        volume=volume, 
                        ventilation_rate = .1, 
                        CO2var = 2,
                        method='Exponential', 
                        freq=freq, useCpp = FALSE), iterations = 100) 
Rtimes <- as.numeric(Rresults$time[[1]]) 
Cpptimes <- as.numeric(Cppresults$time[[1]])
mean(Rtimes)
sd(Rtimes)
mean(Cpptimes)
sd(Cpptimes)
```

Perhaps the environmental CO2 and/or emission rates are not measured. We can use a multi-dimensional optimization method to estimate Q, environmental CO2, and/or emission rates.

(insert equations here)

Let's see how it performs on simulated data:

```{r}
freq=1
volume=500
envCO2=380
Q = .01
temp =25

persondata <- data.frame(
  time = c(0, 8, 8, 17, 24),
  n = c(0, 15, 2,  0, 0),
  age = c(0, 8, 30,  0, 0),
  gender = rep(NA, 5),
  MET = c(NA, 3, 1.8,  NA, NA),
  CO2rate = rep(NA, 5)
)

critpoints = c(3600/freq*8, 3600/freq*12, 3600/freq*13, 3600/freq*17)
simdata <- simulateData(persondata = persondata,
                        volume=volume, 
                        ventilation_rate = Q, 
                        envCO2=envCO2,
                        startCO2=envCO2,
                        CO2var = 0.01,
                        method='Exponential', 
                        freq=freq)

ggplot(simdata, aes(x=time, y=CO2)) + geom_line() + labs(x="Time (hours)", y="CO2 (ppm)") + theme_bw() 

true_emissions <- persondata_to_emission(persondata, temp, freq)
true_emission_rates <- unique(true_emissions$nadj_CO2rate)

# testing with trying to estimate emission rates: doesn't work very well
keep_these <- seq(1, length(simdata$CO2), 60)
test <- estimate_ventilation(freq=60,
                     CO2=simdata$CO2[keep_these],
                     volume=volume,
                     persondata=persondata,
                     init.Q=.5,
                     temp=temp,
                     critpoints=critpoints,
                     method='Newton',
                     max.iter=1000,
                     tol=1e-8,
                     envCO2.init=425, verbose=T,
                     record.steps = F)
test

```

```{r}
freq=1
volume=500
temp =25

persondata <- data.frame(
  time = c(0, 8, 8, 17, 24),
  n = c(0, 15, 2,  0, 0),
  age = c(0, 8, 30,  0, 0),
  gender = rep(NA, 5),
  MET = c(NA, 3, 1.8,  NA, NA),
  CO2rate = rep(NA, 5)
)
freq.est = 60
testing_Newton <- function(Q, envCO2, var, init.Q, init.envCO2){
  sim <- simulateData(persondata = persondata,
                        volume=volume, 
                        ventilation_rate = Q, 
                        CO2var = var,
                        method='Exponential', 
                        freq=freq,
                      envCO2=envCO2)
  index = seq(1, length(sim$CO2), freq.est)
  CO2 = sim$CO2[index]
  out <- tryCatch({
    estimate_ventilation(freq=freq.est, 
                     CO2=CO2, 
                     volume=volume, 
                     envCO2.init=init.envCO2, 
                     init.Q=init.Q, 
                     temp=temp, 
                     persondata=persondata, 
                     method='Newton', 
                     max.iter=1000, 
                     tol=1e-6)
                  },
           error= function(e) return(list('ventilation' = NA, 'envCO2' = NA, 'iter' = NA, 'convergence' = NA)))
  return(c(Q.est = out$ventilation, envCO2.est = out$envCO2, iter = out$iter, convergence = out$convergence, Q.true = Q, envCO2.true = envCO2, var = var, Q.init = init.Q, envCO2.init = init.envCO2))
}

Q = rep(seq(0.01, .5, length=5)) 
envCO2 = rep(c(385, 400, 415))
Q.init = seq(.1, 1, length=10)
envCO2.init = c(375, 400, 425)
var = .01
test_cases = matrix(NA, nrow = 0, ncol = 4)
for(i in 1:length(Q)){
  for(j in 1:length(envCO2)){
    for(k in 1:length(Q.init)){
      for(l in 1:length(envCO2.init)){
        params <- c(Q[i], envCO2[j], Q.init[k], envCO2.init[l])
        test_cases <- rbind(test_cases, params)
      }
    }
  }
}


out_results = matrix(NA, nrow = nrow(test_cases), ncol = 9)
colnames(out_results) = c("Q.est", "envCO2.est", "iter", "convergence", "Q.true", "envCO2.true", "var", "Q.init", "envCO2.init")
for(i in 1:nrow(test_cases)){
  #print(paste("Running test case", i, "of", nrow(test_cases), "test cases."))
  out_results[i,] = testing_Newton(test_cases[i, 1], test_cases[i, 2], var=.01, init.Q=test_cases[i, 3], init.envCO2=test_cases[i, 4])
  #print(testing_Newton(test_cases[i, 1], test_cases[i, 2], var=.01, init.Q=.5, init.envCO2=425))
}

out_results <- as.data.frame(out_results)
out_results <- out_results %>% mutate(rel.error.Q = abs(Q.est-Q.true)/Q.true) %>% mutate(rel.error.env = abs(envCO2.est-envCO2.true)/envCO2.true)


summary <- out_results %>% group_by(Q.true, envCO2.true) %>% summarize(mean.Q.est = mean(na.omit(Q.est)), mean.iter = mean(na.omit(iter)), mean.convergence = mean(na.omit(convergence)), mean.rel.error.Q = mean(na.omit(rel.error.Q)), sd.rel.error.Q = sd(na.omit(rel.error.Q)), mean.rel.error.env = mean(na.omit(rel.error.env)), sd.rel.error.env = sd(na.omit(rel.error.env)),NAs = sum(is.na(iter)))

library(tidyr)
summary <- summary %>% pivot_longer(cols = c(mean.rel.error.Q, mean.rel.error.env), names_to = "parameter", values_to = "mean.rel.error")
summary <- summary %>% pivot_longer(cols = c(sd.rel.error.Q, sd.rel.error.env), names_to = "parameter2", values_to = "sd.rel.error")
summary <- summary %>% mutate(parameter = factor(parameter, levels = c("mean.rel.error.Q", "mean.rel.error.env"), labels=c("Ventilation rate", "Environmental CO2")))
summary <- summary %>% mutate(parameter2 = factor(parameter2, levels = c("sd.rel.error.Q", "sd.rel.error.env"), labels=c("Ventilation rate", "Environmental CO2")))
summary <- summary %>% filter(parameter == parameter2)

ggplot(summary, aes(x=Q.true, y=mean.rel.error, color=parameter))+     geom_point() + 
  geom_line() +  
  geom_ribbon(aes(ymin=mean.rel.error-sd.rel.error, ymax=mean.rel.error+sd.rel.error), alpha=.2) + facet_wrap(~paste('Simulated environmental CO2:', envCO2.true))+
  labs(x="True Ventilation Rate (m3/s)", y="Relative error in estimated parameters") + theme_bw()  + labs(color="Parameter")

sum(summary$NAs)/2/450

```
```{r}
# testing on real data
# real data set-up
data = readxl::read_excel("Aranet4 0C1B8_2023-11-15T18_47_56-0500.xlsx")
times = data$`Time(dd/mm/yyyy)`
times = as.POSIXct(times, format = "%d/%m/%Y %I:%M:%S %p")
CO2 = data$`Carbon dioxide(ppm)`
ggplot(data=data.frame(times=times, CO2=CO2), aes(x=times, y=CO2)) + 
  geom_point() + 
  geom_line() + 
  labs(x="Time", y="CO2 (ppm)") + theme_bw() 

ind0 <- which(times == "2023-11-13 18:01:35 EST") # people leave for the day
ind1 <- which(times == "2023-11-14 8:01:36 EST") # people come in
ind2 <- which(times == "2023-11-14 18:01:36 EST") # people leave
ind3 <- which(times == "2023-11-15 8:01:35 EST") # people come in
ind4 <- which(times == "2023-11-15 18:01:35 EST") # people leave

hr0 <- as.numeric(difftime(times[ind0], times[1], units = "hours"))
hr1 <- as.numeric(difftime(times[ind1], times[1], units = "hours"))
hr2 <- as.numeric(difftime(times[ind2], times[1], units = "hours"))
hr3 <- as.numeric(difftime(times[ind3], times[1], units = "hours"))
hr4 <- as.numeric(difftime(times[ind4], times[1], units = "hours"))
end <- as.numeric(difftime(times[length(times)], times[1], units = "hours"))


persondata3 <- data.frame(
  time = c(0, hr0, hr1, hr2, hr3, hr4, end),
  n = c(3, 0, 3, 0, 3, 0, 0))

persondata5 <- data.frame(
  time = c(0, hr0, hr1, hr2, hr3, hr4, end),
  n = c(5, 0, 5, 0, 5, 0, 0))

persondata0 <- data.frame(
  time = c(0, end),
  n = c(0, 0))

volume = .0254^3*321*378*114
envCO2 = 400
temp=25


# averaging data
library(zoo)
CO2_avg <- rollmean(CO2, 15, fill=NA, align="right")
ggplot(data=data.frame(times=times, CO2=CO2_avg), aes(x=times, y=CO2)) + 
  geom_point() + 
  geom_line() + 
  labs(x="Time", y="CO2 (ppm)") + theme_bw()
```

```{r}
# testing methods on real data
# note, ACHPH = 3600Q/Volume

NR_real_3ppl <- estimate_ventilation(freq=120, 
                     CO2=CO2, 
                     volume=volume, 
                     envCO2known=400, 
                     init.Q=1, 
                     temp=temp, 
                     persondata=persondata3, 
                     method='NR', 
                     max.iter=1000, 
                     tol=1e-10)
NR_real_3ppl$est_ventilation*3600/volume

NR_real_5ppl <- estimate_ventilation(freq=120, 
                     CO2=CO2, 
                     volume=volume, 
                     envCO2known=400, 
                     init.Q=1, 
                     temp=temp, 
                     persondata=persondata5, 
                     method='NR', 
                     max.iter=1000, 
                     tol=1e-10)
NR_real_5ppl$est_ventilation*3600/volume

NR_real_0ppl <- estimate_ventilation(freq=120, 
                     CO2=CO2, 
                     volume=volume, 
                     envCO2known=400, 
                     init.Q=1, 
                     temp=temp, 
                     persondata=persondata0, 
                     method='NR', 
                     max.iter=1000, 
                     tol=1e-10)
NR_real_0ppl$est_ventilation*3600/volume

NR_real_3ppl_averaged <- estimate_ventilation(freq=120, 
                     CO2=CO2_avg[!is.na(CO2_avg)], 
                     volume=volume, 
                     envCO2known=400, 
                     init.Q=1, 
                     temp=temp, 
                     persondata=persondata3, 
                     method='NR', 
                     max.iter=1000, 
                     tol=1e-10) 
NR_real_3ppl_averaged$est_ventilation*3600/volume

NR_real_5ppl_averaged <- estimate_ventilation(freq=120, 
                     CO2=CO2_avg[!is.na(CO2_avg)], 
                     volume=volume, 
                     envCO2known=400, 
                     init.Q=1, 
                     temp=temp, 
                     persondata=persondata5, 
                     method='NR', 
                     max.iter=1000, 
                     tol=1e-10)
NR_real_5ppl_averaged$est_ventilation*3600/volume

Newton_real_3ppl <- estimate_ventilation(freq=120, 
                     CO2=CO2, 
                     volume=volume, 
                     envCO2.init=420, 
                     init.Q=NR_real_3ppl$est_ventilation, 
                     temp=temp, 
                     persondata=persondata3, 
                     method='Newton', 
                     max.iter=1000, 
                     tol=10, verbose=T) # increase tolerance to avoid singular Hessian

#still doesn't really work when considering averaged data
Newton_real_3ppl_avg <- estimate_ventilation(freq=120, 
                     CO2=CO2_avg[!is.na(CO2_avg)], 
                     volume=volume, 
                     envCO2.init=420, 
                     init.Q=NR_real_3ppl$est_ventilation, 
                     temp=temp, 
                     persondata=persondata3, 
                     method='Newton', 
                     max.iter=1000, 
                     tol=10, verbose=T)
```




```{r}
# Real data for testing
data = readxl::read_excel("Aranet4 0C1B8_2023-11-15T18_47_56-0500.xlsx")
data2 = readxl::read_excel("Aranet4 17877_2023-11-15T18_49_59-0500.xlsx")
times = data$`Time(dd/mm/yyyy)`
times = as.POSIXct(times, format = "%d/%m/%Y %I:%M:%S %p")
times2 = as.POSIXct(data2$`Time(dd/mm/yyyy)`, format = "%d/%m/%Y %I:%M:%S %p")
CO2 = data$`Carbon dioxide(ppm)`
CO22 = data2$`Carbon dioxide(ppm)`

ggplot(data=data.frame(times=times, CO2=CO2), aes(x=times, y=CO2)) + 
  geom_line()

ggplot(data=data.frame(times=times2, CO2=CO22), aes(x=times, y=CO2)) + 
  geom_line()

periods1 <- exp_periods(times, CO2, quant_cutoff=.9)
periods1$timediff = difftime(periods1$Time.Start, times[1], units='hours')
# remove redundancies
periods1 <- periods1[c(1,3,5,6,7,8,9,10,11,14,15,16),]

periods2 <- exp_periods(times2, CO22, quant_cutoff = .95)
periods2$timediff = difftime(periods2$Time.Start, times2[1], units='hours')
periods2 <- periods2[c(1, 2, 4, 5, 8, 9, 12, 13, 15, 17, 18, 19, 21),]

# fake persondata based on exponential periods
persondata1 <- data.frame(
  time = c(as.numeric(periods1$timediff), as.numeric(difftime(times[length(times)], times[1], unit='hours'))),
  n = c(3*as.numeric(periods1$Type=='Growth'), 0)
)

persondata2 <- data.frame(
  time = c(0, as.numeric(periods2$timediff), as.numeric(difftime(times2[length(times2)], times2[1], unit='hours'))),
  n = c(0, 3*as.numeric(periods2$Type=='Growth'), 0)
)
# update with exponential periods and use n=3

volume = .0254^3*321*378*114
envCO2 = 400
temp=25

# 6 ACH minimum ventilation when occupied

# ACHPH = 3600Q/Volume

realdata <- estimate_ventilation(freq=120, 
                     CO2=CO22, 
                     volume=volume, 
                     envCO2known=400, 
                     init.Q=1, 
                     temp=temp, 
                     persondata=persondata2, 
                     method='NR', 
                     max.iter=1000, 
                     tol=1e-10)

realdata$est_ventilation/volume*3600

realdata_Newton <- estimate_ventilation(freq=120, 
                     CO2=CO22, 
                     volume=volume,
                     init.Q=1, 
                     temp=temp, 
                     persondata=persondata2, 
                     method='Newton', 
                     max.iter=1000, 
                     envCO2.init=400,
                     tol=1e-6, verbose=T)

realdata_Newton$ventilation/volume*3600

ug_to_ppm <- function(ug, temp) {
    ug * 44.01 * 1000 * 101.325 / 8314 / (273.15+temp)
  }
ug_to_ppm(realdata_Newton$envCO2, 25)

realdata_Newton_estE <- estimate_ventilation(freq=120, 
                     CO2=CO22, 
                     volume=volume,
                     init.Q=1, 
                     temp=temp, 
                     critpoints= periods2$Index.Start,
                     method='Newton', 
                     max.iter=1000, 
                     envCO2.init=500,
                     tol=1e-6, verbose=T)
realdata_Newton_estE$ventilation*3600/volume
ug_to_ppm(realdata_Newton_estE$envCO2, 25)
realdata_Newton_estE$E

ggplot(data=data.frame(times=times, CO2=CO2, est_E = c(realdata_Newton_estE$E,0)), aes(x=times, y=CO2, color=est_E)) + 
  geom_line() + 
  geom_smooth()

loess_curve <- loess(CO2 ~as.numeric(times), span=.20, degree=2 )
plot(loess_curve$x, loess_curve$fitted)
plot(as.numeric(times), CO2)

# extract locations of local minima and maxima in loess curve fitted values
maxima <- which(diff(sign(diff(loess_curve$fitted)))==-2)+1
minima <- which(diff(sign(diff(loess_curve$fitted)))==2)+1
critpoints <- c(maxima, minima)

ggplot(data=data.frame(times=times, CO2=CO2), aes(x=times, y=CO2)) + 
  geom_line() + 
  geom_smooth(method='loess', span=.2) + 
  geom_vline(xintercept=times[critpoints], color='red')

ggplot(data=data.frame(times=times, CO2=loess_curve$fitted), aes(x=times, y=CO2)) + 
  geom_line() + 
  geom_vline(xintercept=times[critpoints], color='red')

times3 = difftime(times2[sort(critpoints)], times2[1], units='hours')
persondata3 <- data.frame(
  time = c(0, as.numeric(times3), as.numeric(difftime(times2[length(times2)], times2[1], unit='hours'))),
  n = c(3, 0, 3, 3, 0, 0, 3, 0, 3, 0, 3, 0,0)
)


realdata_Newton_estE <- estimate_ventilation(freq=120, 
                     CO2=CO22, 
                     volume=volume,
                     init.Q=1, 
                     temp=temp, 
                     persondata=persondata3,
                     method='Newton', 
                     max.iter=1000, 
                     envCO2.init=400,
                     tol=1e-6, verbose=T)
realdata_Newton_estE$ventilation*3600/volume
ug_to_ppm(realdata_Newton_estE$envCO2, 25)
realdata_Newton_estE$E
```
